---
title: "first work"
output: html_document
---
``` {r setup, include=FALSE}
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Descriptive plot to get to know data
### Tomb parameter
Clean the data: 
- delete single entries for T46, T56, T62
- create entries for them as 1 tomb but with difference MNIs (as project partner wants)

``` {r tp}
tomb_parameter <- read_excel("Data/raw/Statistics project LMU_tomb parameters_Irene Hoegner_new.xlsx", range = "A2:N12")
combined_tombs <- tomb_parameter |> slice(rep(10, 4)) |> mutate(
  "MNI (minimum number of individuals)" = c(250, 300, 350, 400),
  "percentage of successful samples" = `TWIST capture samples included in kinship analysis` / `MNI (minimum number of individuals)`,
  `...1` = paste("Elateia T 46,56,62 (MNI ", c(250, 300, 350, 400), ")", sep = ""))
tomb_parameter <- rbind(tomb_parameter, combined_tombs)
tomb_parameter <- tomb_parameter |> rename(Tomb = 1, MNI = 2, `analysed individuals` = 12) |>
  slice(-c(4, 7, 9, 10)) |> mutate(MNI = as.integer(MNI))
tomb_parameter
```

The important info in this dataset: "MNI" and "after merging samples from same individual" (**minimum # individuals** vs **actual successful analysed # individual**), since if we don't consider the ratio between them, the result could be bias.

Therefore I create plots of absolute and relative relation of those numbers for every tombs:

``` {r plot tp}
# Absolute plot
df_long <- pivot_longer(tomb_parameter, cols = c(MNI, `analysed individuals`),
                        names_to = "Type", values_to = "Count")
ggplot(df_long, aes(x = Tomb, y = Count, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge")

```
