---
title: "Denominator + Nonsensical covariates combination + Network"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
source(here::here("environmentSetUp.R"))
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(igraph)
library(ggraph)
library(purrr)
```
## Denominator
**Question:** MNI cannot be interpreted as the population unless sampling was random, but it wasn’t. So what should be the denominator (First selection layer)?

**Before:** Using MNI to calculate the maximum number of pairs.

Problem: 

- Numerator = # observed related pairs only come from the **analysed individuals**
- Using MNI creates sample selection bias, since we assume the population (MNI) have the similar kinship patterns with the analysed subset (the analysed subset is not randomly selected)

$\Rightarrow$ **# Analysed individuals** as denominator *(= "after merging samples from same individual", column L in tomb_parameter)*


## Nonsensical covariate combinations
**Question:** Check for nonsensical covariate combinations to see if they are "even modelable" before running models.

### Check for nonsensical covariate combinations


``` {r p, echo=FALSE} 
# Pre-processing data
individual_metadata <- og_individual_metadata |> 
  filter(Site != "Lapoutsi", Tomb %in% c("Amfissa tholos", "Tomb 31", "Tomb 36", "Tomb 46", "Tomb 50", "Tomb 56", "Tomb 62", "Tomb 67")) |>
  mutate(`Age estimation` = if_else(`Age estimation` == "subadult?", "subadult", `Age estimation`),
         `Age estimation` = if_else(`Age estimation` == "adolescent or young adult", "adult or undefined", `Age estimation`))

tomb_group_rows <- individual_metadata |>
  filter(Tomb %in% c("Tomb 46", "Tomb 56", "Tomb 62")) |>
  mutate(Tomb = "Tomb 46 56 62")

individual_metadata <- bind_rows(individual_metadata, tomb_group_rows)
```

#### Skeletal element
**Absolute number of skeletal element in every tomb**
``` {r 1, echo=FALSE}
individual_metadata |>
  count(Tomb, `Skeletal element`) |>
  group_by(Tomb) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = `Skeletal element`, y = prop, fill = `Skeletal element`)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Tomb) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  ylab("Proportion of Skeletal Elements") +
  xlab("Skeletal Element")
```

**Absolute number of skeletal element as Sex in every tomb**
``` {r 2, echo=FALSE}
individual_metadata |>
  count(Tomb, `Skeletal element`, Sex) |>
  ggplot(aes(x = `Skeletal element`, y = n, fill = Sex)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ Tomb) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Count") +
  xlab("Skeletal Element")
```

**Relative number of skeletal element as Sex in every tomb**
``` {r 3, echo=FALSE}
individual_metadata |>
  count(Tomb, `Skeletal element`, Sex) |>
  group_by(Tomb, `Skeletal element`) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = `Skeletal element`, y = prop, fill = Sex)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ Tomb) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion of Sex within Skeletal Element") +
  xlab("Skeletal Element")
```


**Absolute number of skeletal element as Age in every tomb**
``` {r 4, echo=FALSE}
individual_metadata |>
  count(Tomb, `Skeletal element`, `Age estimation`) |>
  ggplot(aes(x = `Skeletal element`, y = n, fill = `Age estimation`)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ Tomb) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Count") +
  xlab("Skeletal Element")
```

**Relative number of skeletal element as Age in every tomb**
``` {r 5, echo=FALSE}
individual_metadata |>
  count(Tomb, `Skeletal element`, `Age estimation`) |>
  group_by(Tomb, `Skeletal element`) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = `Skeletal element`, y = prop, fill = `Age estimation`)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ Tomb) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion of Sex within Skeletal Element") +
  xlab("Skeletal Element")
```



**Absolute and Relative number of Sex as Age in every tomb**

``` {r 6, echo=FALSE}
individual_metadata |>
  count(Tomb, `Age estimation`, Sex) |>
  ggplot(aes(x = `Age estimation`, y = n, fill = Sex)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ Tomb) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Count") +
  xlab("Age Estimation")


individual_metadata |>
  count(Tomb, `Age estimation`, Sex) |>
  group_by(Tomb) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = `Age estimation`, y = prop, fill = Sex)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ Tomb) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion of Sex") +
  xlab("Age Estimation")

```


#### Problems:
Sparse data. In some tombs:

- lack certain skeletal elements
- some skeletal elements only appear in one sex or age

$\Rightarrow$ Problem: 

- Unstable model, effect of the missing skeletal element can not be estimated for that tomb
- Collinearity: eg. Tomb 67 has only female phalanx $\Rightarrow$ perfect collinearity between skeletal element & sex?

Previous problem with the GLM logistic model:

- With-in tomb correlation (same preservation environment,...)
- Too many parameters with too few observation (tomb-level data set)

**Solution: ** Mixed logistic modelling on individual-level data. Especially GLMM can handle categorical predictors with some levels missing in some cluster. It also assumes MAR, which is very realistic in our case.

Source:

https://stats.stackexchange.com/questions/626546/how-does-lme4-handle-participants-who-only-have-data-for-only-one-level-of-a-cat

https://www.sciencedirect.com/science/article/pii/S0015028224002371



## Network visualization
### All tombs together
``` {r kr 2, echo=FALSE}
kinship_result <- read_excel("Data/raw/Statistics project LMU_READv2 kinship results_Irene Hoegner_corr.xlsx")
og_individual_metadata <- read_excel("Data/raw/Statistics project LMU_individuals metadata_Irene Hoegner.xlsx")

kinship_result <- kinship_result |>
  mutate(Rel = if_else(Low_data == "yes", "Uncertain", Rel, missing = Rel),
  "Individual1 ID" = substr(Ind1, 1, 6),
  "Individual2 ID" = substr(Ind2, 1, 6)) |>
  left_join(og_individual_metadata, by = c("Individual1 ID" = "Individual ID")) |>
  rename(Tomb1 = Tomb) |>
  left_join(og_individual_metadata, by = c("Individual2 ID" = "Individual ID")) |>
  rename(Tomb2 = Tomb) |> 
  select(names(kinship_result), Tomb1, Tomb2) |>
  mutate(
    "Individual1 ID" = substr(Ind1, 1, 6),
    "Individual2 ID" = substr(Ind2, 1, 6)
  ) |>
  filter(!startsWith(`Ind1`, "LPS") & !startsWith(`Ind2`, "LPS"),
         Tomb1 %in% c("Amfissa tholos", "Tomb 31", "Tomb 36", "Tomb 46", "Tomb 50", "Tomb 56", "Tomb 62", "Tomb 67"),
         Tomb2 %in% c("Amfissa tholos", "Tomb 31", "Tomb 36", "Tomb 46", "Tomb 50", "Tomb 56", "Tomb 62", "Tomb 67")) 

grouped_tombs <- c("Tomb 46", "Tomb 56", "Tomb 62")
group_name <- "Tomb 46 56 62"

kinship_group_pairs <- kinship_result |>
  filter(Tomb1 %in% grouped_tombs | Tomb2 %in% grouped_tombs) |>
  
  # Rename the Tomb columns in the DUPLICATE rows to reflect the new group
  mutate(
    Tomb1 = ifelse(Tomb1 %in% grouped_tombs, group_name, Tomb1),
    Tomb2 = ifelse(Tomb2 %in% grouped_tombs, group_name, Tomb2)
  )

kinship_result <- bind_rows(kinship_result, kinship_group_pairs)

# COUNT PAIRS
count_pairs <- kinship_result |> group_by(Tomb1, Tomb2, Rel) |> summarize(num_pairs = n())
internal_relation <- count_pairs[count_pairs$Tomb1 == count_pairs$Tomb2, ]  # 2 Inds inside same tomb



# TEST NETWORK VISUALIZATION
network_kr <- kinship_result %>% 
  mutate(
    Related = case_when(
      Rel %in% c("First Degree", "IdenticalTwins/SameIndividual", "Second Degree", "Third Degree") ~ TRUE,
      Rel == "Uncertain" ~ FALSE,
      Rel %in% c("Unrelated", "Unrelated/Consistent with Third Degree") ~ NA
    )
  ) %>% 
  filter(Related == TRUE)

edges <- network_kr %>% 
  select(`Individual1 ID`, `Individual2 ID`, Tomb1, Tomb2)

# Create nodes using both columns and their tombs
nodes <- edges %>%
  transmute(name = `Individual1 ID`, Tomb = Tomb1) %>%
  bind_rows(edges %>% transmute(name = `Individual2 ID`, Tomb = Tomb2)) %>%
  distinct(name, .keep_all = TRUE)

# Create graph
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)

# Plot
ggraph(g, layout = "fr") + 
  geom_edge_link(color = "grey70") +
  geom_node_point(aes(color = Tomb), size = 3) +
  geom_node_text(aes(label = name), vjust = -1, size = 2) +
  theme_void() +
  scale_color_brewer(palette = "Set2")
```


### Separated tombs
``` {r sep, echo=FALSE}
tombs <- unique(nodes$Tomb)

plots <- map(tombs, function(t) {
  
  # select edges where either individual belongs to tomb t
  edges_t <- edges %>%
    filter(Tomb1 == t | Tomb2 == t)
  
  # nodes involved
  nodes_t <- nodes %>%
    filter(name %in% unique(c(edges_t$`Individual1 ID`, edges_t$`Individual2 ID`)))
  
  # graph
  g_t <- graph_from_data_frame(d = edges_t, vertices = nodes_t, directed = FALSE)
  
  # plot
  ggraph(g_t, layout = "fr") + 
    geom_edge_link(color = "grey70") +
    geom_node_point(aes(color = Tomb), size = 3) +
    geom_node_text(aes(label = name), vjust = -1, size = 2) +
    theme_void() +
    ggtitle(paste(t)) +
    scale_color_brewer(palette = "Set2")
})


for(p in plots) print(p)


```



#### Grouped tomb 46 56 62 version
``` {r 3 tombs, echo=FALSE}
kinship_result_grouped <- kinship_result %>%
  mutate(
    Tomb1 = ifelse(Tomb1 %in% grouped_tombs, group_name, Tomb1),
    Tomb2 = ifelse(Tomb2 %in% grouped_tombs, group_name, Tomb2)
  )

network_grouped <- kinship_result_grouped %>%
  mutate(
    Related = case_when(
      Rel %in% c("First Degree", "IdenticalTwins/SameIndividual", "Second Degree", "Third Degree") ~ TRUE,
      Rel == "Uncertain" ~ FALSE,
      TRUE ~ NA
    )
  ) %>%
  filter(Related == TRUE)

edges_g <- network_grouped %>%
  select(`Individual1 ID`, `Individual2 ID`, Tomb1, Tomb2)

nodes_g <- edges_g %>%
  transmute(name = `Individual1 ID`, Tomb = Tomb1) %>%
  bind_rows(edges_g %>% transmute(name = `Individual2 ID`, Tomb = Tomb2)) %>%
  distinct(name, .keep_all = TRUE)

nodes_g$Tomb <- droplevels(factor(nodes_g$Tomb))

g_g <- graph_from_data_frame(d = edges_g, vertices = nodes_g, directed = FALSE)

ggraph(g_g, layout = "fr") +
  geom_edge_link(color = "grey70") +
  geom_node_point(aes(color = Tomb), size = 3) +
  geom_node_text(aes(label = name), vjust = -1, size = 2) +
  theme_void() +
  ggtitle("All tombs with grouped Tomb 46 56 62") +
  scale_color_brewer(palette = "Set2")
```

#### Tomb 46 56 62 only

``` {r only, echo=FALSE}
group_name <- "Tomb 46 56 62"

network_grouped <- kinship_result_grouped %>%
  # keep only pairs where at least one side is the grouped tomb
  filter(Tomb1 == group_name | Tomb2 == group_name) %>%
  
  mutate(
    Related = case_when(
      Rel %in% c("First Degree", "IdenticalTwins/SameIndividual", 
                 "Second Degree", "Third Degree") ~ TRUE,
      Rel == "Uncertain" ~ FALSE,
      TRUE ~ NA
    )
  ) %>%
  filter(Related == TRUE)

edges_g <- network_grouped %>%
  select(`Individual1 ID`, `Individual2 ID`, Tomb1, Tomb2)

nodes_g <- edges_g %>%
  transmute(name = `Individual1 ID`, Tomb = Tomb1) %>%
  bind_rows(edges_g %>% transmute(name = `Individual2 ID`, Tomb = Tomb2)) %>%
  distinct(name, .keep_all = TRUE)

nodes_g$Tomb <- droplevels(factor(nodes_g$Tomb))

g_g <- graph_from_data_frame(d = edges_g, vertices = nodes_g, directed = FALSE)

ggraph(g_g, layout = "fr") +
  geom_edge_link(color = "grey70") +
  geom_node_point(aes(color = Tomb), size = 3) +
  geom_node_text(aes(label = name), vjust = -1, size = 2) +
  theme_void() +
  ggtitle("Network: Only pairs involving grouped Tomb 46 56 62") +
  scale_color_brewer(palette = "Set2")

```


**Comment:**

- Different ratio in whether the related pairs or in the same tombs or in different tombs
- But just by looking, some tombs really point out that they tends to have internal relatedness more than random relatedness to different tombs.











