---
title: "Denominator + Nonsensical covariates combination + Network"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
source(here::here("environmentSetUp.R"))
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(igraph)
library(ggraph)
library(purrr)
```

## Denominator

**Question:** MNI cannot be interpreted as the population unless sampling was random, but it wasn’t. So what should be the denominator (First selection layer)?

**Before:** Using MNI to calculate the maximum number of pairs.

Problem:

-   Numerator = \# observed related pairs only come from the **analysed individuals**
-   Using MNI creates sample selection bias, since we assume the population (MNI) have the similar kinship patterns with the analysed subset (the analysed subset is not randomly selected)

$\Rightarrow$ **\# Analysed individuals** as denominator *(= "after merging samples from same individual", column L in tomb_parameter)*

## Nonsensical covariate combinations

**Question:** Check for nonsensical covariate combinations to see if they are "even modelable" before running models.

### Check for nonsensical covariate combinations

```{r p, echo=FALSE}
# Pre-processing data
individual_metadata <- og_individual_metadata |> 
  filter(
    site != "Lapoutsi",
    tomb %in% c("Amfissa tholos", "Tomb 31", "Tomb 36", 
                "Tomb 46", "Tomb 50", "Tomb 56", "Tomb 62", "Tomb 67")
  ) |>
  mutate(
    age_estimation = if_else(age_estimation == "subadult?", "subadult", age_estimation),
    age_estimation = if_else(age_estimation == "adolescent or young adult", 
                             "adult or undefined", age_estimation)
  )

tomb_group_rows <- individual_metadata |>
  filter(tomb %in% c("Tomb 46", "Tomb 56", "Tomb 62")) |>
  mutate(tomb = "Tomb 46 56 62")

individual_metadata <- bind_rows(individual_metadata, tomb_group_rows)

```

#### Skeletal element

**Absolute number of skeletal element in every tomb**

```{r 1, echo=FALSE}
individual_metadata |>
  count(tomb, skeletal_element) |>
  group_by(tomb) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = skeletal_element, y = prop, fill = skeletal_element)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ tomb) +
  scale_y_continuous(labels = scales::percent) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion of Skeletal Elements") +
  xlab("Skeletal Element")
```

**Absolute number of skeletal element as Sex in every tomb**

```{r 2, echo=FALSE}
individual_metadata |>
  count(tomb, skeletal_element, sex) |>
  ggplot(aes(x = skeletal_element, y = n, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ tomb) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Count") +
  xlab("Skeletal Element")
```

**Relative number of skeletal element as Sex in every tomb**

```{r 3, echo=FALSE}
individual_metadata |>
  count(tomb, skeletal_element, sex) |>
  group_by(tomb, skeletal_element) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = skeletal_element, y = prop, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ tomb) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion of Sex within Skeletal Element") +
  xlab("Skeletal Element")
```

**Absolute number of skeletal element as Age in every tomb**

```{r 4, echo=FALSE}
individual_metadata |>
  count(tomb, skeletal_element, age_estimation) |>
  ggplot(aes(x = skeletal_element, y = n, fill = age_estimation)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ tomb) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Count") +
  xlab("Skeletal Element")
```

**Relative number of skeletal element as Age in every tomb**

```{r 5, echo=FALSE}
individual_metadata |>
  count(tomb, skeletal_element, age_estimation) |>
  group_by(tomb, skeletal_element) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = skeletal_element, y = prop, fill = age_estimation)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ tomb) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion of Age Estimation within Skeletal Element") +
  xlab("Skeletal Element")
```

**Absolute and Relative number of Sex as Age in every tomb**

```{r 6, echo=FALSE}
individual_metadata |>
  count(tomb, age_estimation, sex) |>
  ggplot(aes(x = age_estimation, y = n, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ tomb) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Count") +
  xlab("Age Estimation")


individual_metadata |>
  count(tomb, age_estimation, sex) |>
  group_by(tomb) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = age_estimation, y = prop, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ tomb) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion of Sex") +
  xlab("Age Estimation")

```

#### Problems:

Sparse data. In some tombs:

-   lack certain skeletal elements
-   some skeletal elements only appear in one sex or age

$\Rightarrow$ Problem:

-   Unstable model, effect of the missing skeletal element can not be estimated for that tomb
-   Collinearity: eg. Tomb 67 has only female phalanx $\Rightarrow$ perfect collinearity between skeletal element & sex?

Previous problem with the GLM logistic model:

-   With-in tomb correlation (same preservation environment,...)
-   Too many parameters with too few observation (tomb-level data set)

**Solution:** Mixed logistic modelling on individual-level data. Especially GLMM can handle categorical predictors with some levels missing in some cluster. It also assumes MAR, which is very realistic in our case.

Source:

<https://stats.stackexchange.com/questions/626546/how-does-lme4-handle-participants-who-only-have-data-for-only-one-level-of-a-cat>

<https://www.sciencedirect.com/science/article/pii/S0015028224002371>

::: {style="color: red;"}
#### Yichen's feedback:

Conclusion is fine, but visualisation needs a smarter approach for presentation.

Correction idea:

Use heatmap to show exactly where the data is missing to justify the use of GLMM. Like instead of showing a bunch of bar charts, we show the audience the zeros directly in a heatmap for easier comparison.
:::

##### Absolute number of skeletal element as Sex in every tomb

```{r hm abs sex, echo=FALSE}
# Create Tomb × Skeletal Element × Sex count matrix
heat_df <- individual_metadata %>%
  count(tomb, sex, skeletal_element) %>%
  complete(tomb, sex, skeletal_element, fill = list(n = 0))

# Plot
ggplot(heat_df, aes(x = skeletal_element, y = tomb, fill = n)) +
  geom_tile(color = "black") +
  facet_wrap(~ sex) +
  scale_fill_gradient(
    name = "Count",
    low = "white",
    high = "darkblue"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    x = "Skeletal Element",
    y = "Tomb"
  )

```

##### Relative number of skeletal element as Sex in every tomb

```{r hm rel sex, echo=FALSE}
# Count raw numbers and compute relative proportions
heat_df <- individual_metadata %>%
  count(tomb, sex, skeletal_element) %>%
  complete(tomb, sex, skeletal_element, fill = list(n = 0)) %>%
  group_by(tomb, sex) %>%
  mutate(prop = n / sum(n)) %>%   # <-- relative share per tomb
  ungroup()

# Plot relative proportions
ggplot(heat_df, aes(x = skeletal_element, y = tomb, fill = prop)) +
  geom_tile(color = "black") +
  facet_wrap(~ sex) +
  scale_fill_gradient(
    name = "Proportion",
    low = "white",
    high = "darkblue",
    labels = scales::percent_format(accuracy = 1)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Swiss Cheese Heatmap (Relative): Skeletal Elements by Tomb",
    x = "Skeletal Element",
    y = "Tomb"
  )
```

##### Absolute number of skeletal element as Age in every tomb

```{r abs age, echo = FALSE}
# Create Tomb × Skeletal Element × Age count matrix
heat_df <- individual_metadata %>%
  count(tomb, age_estimation, skeletal_element) %>%
  complete(tomb, age_estimation, skeletal_element, fill = list(n = 0))

# Plot
ggplot(heat_df, aes(x = skeletal_element, y = tomb, fill = n)) +
  geom_tile(color = "black") +
  facet_wrap(~ age_estimation) +
  scale_fill_gradient(
    name = "Count",
    low = "white",
    high = "darkblue"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    x = "Skeletal Element",
    y = "Tomb"
  )


```

##### Relative number of skeletal element as Age in every tomb

```{r hm rel age, echo=FALSE}
# Count raw numbers and calculate proportions
heat_df <- individual_metadata %>%
  count(tomb, age_estimation, skeletal_element) %>%
  complete(tomb, age_estimation, skeletal_element, fill = list(n = 0)) %>%
  group_by(tomb, age_estimation) %>%
  mutate(prop = n / sum(n)) %>%   # relative share per tomb
  ungroup()

# Plot relative proportions
ggplot(heat_df, aes(x = skeletal_element, y = tomb, fill = prop)) +
  geom_tile(color = "black") +
  facet_wrap(~ age_estimation) +
  scale_fill_gradient(
    name = "Proportion",
    low = "white",
    high = "darkblue",
    labels = scales::percent_format(accuracy = 1)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    x = "Skeletal Element",
    y = "Tomb"
  )

```

##### Absolute number of Age as Sex in every tomb

```{r abs age sex, echo = FALSE}
# Create Tomb × Sex × Age count matrix
heat_df <- individual_metadata %>%
  count(tomb, age_estimation, sex) %>%
  complete(tomb, age_estimation, sex, fill = list(n = 0))

# Plot
ggplot(heat_df, aes(x = sex, y = tomb, fill = n)) +
  geom_tile(color = "black") +
  facet_wrap(~ age_estimation) +
  scale_fill_gradient(
    name = "Count",
    low = "white",
    high = "darkblue"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Sex",
    y = "Tomb"
  )

```

##### Relative number of Sex as Age in every tomb

```{r hm rel sex age, echo=FALSE}
# Count raw numbers
heat_df <- individual_metadata %>%
  count(tomb, age_estimation, sex) %>%
  complete(tomb, age_estimation, sex, fill = list(n = 0)) %>%
  group_by(tomb, age_estimation) %>%
  mutate(prop = n / sum(n)) %>%   # <-- relative share per tomb
  ungroup()

# Plot relative proportions
ggplot(heat_df, aes(x = sex, y = tomb, fill = prop)) +
  geom_tile(color = "black") +
  facet_wrap(~ age_estimation) +
  scale_fill_gradient(
    name = "Proportion",
    low = "white",
    high = "darkblue",
    labels = scales::percent_format(accuracy = 1)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Sex",
    y = "Tomb"
  )

```

##### Notes:

Will only use relative plots

## Network visualization

### All tombs together

```{r kr 2, echo=FALSE}
kinship_result <- og_kinship_result %>%
  mutate(
    individual1_id = substr(ind1, 1, 6),
    individual2_id = substr(ind2, 1, 6),
    rel = if_else(low_data == "yes", "Uncertain", rel, missing = rel)
  ) %>%
  
  # join tomb1
  left_join(
    individual_metadata %>% select(individual_id, tomb),
    by = c("individual1_id" = "individual_id")
  ) %>%
  rename(tomb1 = tomb) %>%
  
  # join tomb2
  left_join(
    individual_metadata %>% select(individual_id, tomb),
    by = c("individual2_id" = "individual_id")
  ) %>%
  rename(tomb2 = tomb) %>%
  
  # filter out LPS and invalid tombs
  filter(
    !startsWith(ind1, "LPS"),
    !startsWith(ind2, "LPS"),
    tomb1 %in% c("Amfissa tholos","Tomb 31","Tomb 36","Tomb 46","Tomb 50","Tomb 56","Tomb 62","Tomb 67"),
    tomb2 %in% c("Amfissa tholos","Tomb 31","Tomb 36","Tomb 46","Tomb 50","Tomb 56","Tomb 62","Tomb 67")
  )



# ---- GROUP TOMBS ----

grouped_tombs <- c("Tomb 46", "Tomb 56", "Tomb 62")
group_name <- "Tomb 46 56 62"

kinship_group_pairs <- kinship_result %>%
  filter(tomb1 %in% grouped_tombs | tomb2 %in% grouped_tombs) %>%
  mutate(
    tomb1 = ifelse(tomb1 %in% grouped_tombs, group_name, tomb1),
    tomb2 = ifelse(tomb2 %in% grouped_tombs, group_name, tomb2)
  )

kinship_result <- bind_rows(kinship_result, kinship_group_pairs)



# ---- COUNT PAIRS ----

count_pairs <- kinship_result %>%
  group_by(tomb1, tomb2, rel) %>%
  summarize(num_pairs = n(), .groups = "drop")

internal_relation <- count_pairs[count_pairs$tomb1 == count_pairs$tomb2, ]


# ---- NETWORK VISUALIZATION ----

network_kr <- kinship_result %>% 
  mutate(
    related = case_when(
      rel %in% c("First Degree", "IdenticalTwins/SameIndividual", "Second Degree", "Third Degree") ~ TRUE,
      rel == "Uncertain" ~ FALSE,
      rel %in% c("Unrelated", "Unrelated/Consistent with Third Degree") ~ NA
    )
  ) %>%
  filter(related == TRUE)

edges <- network_kr %>% 
  select(individual1_id, individual2_id, tomb1, tomb2)

nodes <- edges %>%
  transmute(name = individual1_id, tomb = tomb1) %>%
  bind_rows(edges %>% transmute(name = individual2_id, tomb = tomb2)) %>%
  distinct(name, .keep_all = TRUE)

g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)

ggraph(g, layout = "fr") + 
  geom_edge_link(color = "grey70") +
  geom_node_point(aes(color = tomb), size = 3) +
  geom_node_text(aes(label = name), vjust = -1, size = 2) +
  theme_void() +
  scale_color_brewer(palette = "Set2")
```

### With kinship degrees

```{r, echo=FALSE}
# --- Filter TRUE relationships and assign weights manually for plotting ---
network_kr <- kinship_result %>%
  filter(rel %in% c("First Degree", "IdenticalTwins/SameIndividual", "Second Degree", "Third Degree")) %>%
  mutate(
    # Assign plotting widths manually: 1st thickest, 2nd medium, 3rd thinnest
    edge_width = case_when(
      rel %in% c("First Degree", "IdenticalTwins/SameIndividual") ~ 2.0,  # 1st degree
      rel == "Second Degree" ~ 1.2,                                        # 2nd degree
      rel == "Third Degree" ~ 0.6                                          # 3rd degree
    ),
    # Factor for legend ordering
    degree_label = factor(rel,
                          levels = c("First Degree", "Second Degree", "Third Degree"),
                          labels = c("1st Degree", "2nd Degree", "3rd Degree"))
  )

# --- Create edges and nodes ---
edges <- network_kr %>%
  select(individual1_id, individual2_id, tomb1, tomb2, edge_width, degree_label)

nodes <- edges %>%
  transmute(name = individual1_id, tomb = tomb1) %>%
  bind_rows(edges %>% transmute(name = individual2_id, tomb = tomb2)) %>%
  distinct(name, .keep_all = TRUE)

# --- Build graph ---
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)

ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = edge_width, color = degree_label)) +  # show degree in color legend
  geom_node_point(aes(color = tomb), size = 3) +
  geom_node_text(aes(label = name), vjust = -1, size = 2) +
  scale_edge_width(range = c(0.5, 2), guide = "none") +  # widths already set manually
  scale_edge_color_manual(
    values = c("1st Degree" = "black", "2nd Degree" = "grey40", "3rd Degree" = "grey70"),
    name = "Kinship degree"
  ) +
  scale_color_brewer(palette = "Set2") +
  theme_void()

```

### Separated tombs

```{r sep, echo=FALSE}

# list of unique tombs
tombs <- unique(nodes$tomb)

plots <- map(tombs, function(t) {
  
  # select edges where either individual belongs to tomb t
  edges_t <- edges %>%
    filter(tomb1 == t | tomb2 == t)
  
  # nodes involved
  nodes_t <- nodes %>%
    filter(name %in% unique(c(edges_t$individual1_id, edges_t$individual2_id)))
  
  # graph
  g_t <- graph_from_data_frame(d = edges_t, vertices = nodes_t, directed = FALSE)
  
  # plot
  ggraph(g_t, layout = "fr") + 
    geom_edge_link(color = "grey70") +
    geom_node_point(aes(color = tomb), size = 3) +
    geom_node_text(aes(label = name), vjust = -1, size = 2) +
    theme_void() +
    ggtitle(paste(t)) +
    scale_color_brewer(palette = "Set2")
})

# print all plots
for(p in plots) print(p)


```

### Separated tombs with kinship degrees

tombs \<- unique(nodes\$tomb) \# lowercase tomb

```{r, echo=FALSE}
plots <- map(tombs, function(t) {
  
  # select edges where either individual belongs to tomb t
  edges_t <- edges %>%
    filter(tomb1 == t | tomb2 == t)
  
  # nodes involved
  nodes_t <- nodes %>%
    filter(name %in% unique(c(edges_t$individual1_id, edges_t$individual2_id)))
  
  # graph
  g_t <- graph_from_data_frame(d = edges_t, vertices = nodes_t, directed = FALSE)
  
  # plot
  ggraph(g_t, layout = "fr") +
    geom_edge_link(aes(width = edge_width, color = degree_label)) +  # kinship degrees
    geom_node_point(aes(color = tomb), size = 3) +
    geom_node_text(aes(label = name), vjust = -1, size = 2) +
    scale_edge_width(range = c(0.5, 2), guide = "none") +
    scale_edge_color_manual(
      values = c("1st Degree" = "black", "2nd Degree" = "grey40", "3rd Degree" = "grey70"),
      name = "Kinship degree"
    ) +
    scale_color_brewer(palette = "Set2") +
    ggtitle(t) +
    theme_void()
})

for(p in plots) print(p)
```

#### Grouped tomb 46 56 62 version

```{r 3 tombs, echo=FALSE}
# Grouped tombs
kinship_result_grouped <- kinship_result %>%
  mutate(
    tomb1 = ifelse(tomb1 %in% grouped_tombs, group_name, tomb1),
    tomb2 = ifelse(tomb2 %in% grouped_tombs, group_name, tomb2)
  )

# Keep only related pairs
network_grouped <- kinship_result_grouped %>%
  mutate(
    related = case_when(
      rel %in% c("First Degree", "IdenticalTwins/SameIndividual", "Second Degree", "Third Degree") ~ TRUE,
      rel == "Uncertain" ~ FALSE,
      TRUE ~ NA
    )
  ) %>%
  filter(related == TRUE)

# Create edges
edges_g <- network_grouped %>%
  select(individual1_id, individual2_id, tomb1, tomb2)

# Create nodes
nodes_g <- edges_g %>%
  transmute(name = individual1_id, tomb = tomb1) %>%
  bind_rows(edges_g %>% transmute(name = individual2_id, tomb = tomb2)) %>%
  distinct(name, .keep_all = TRUE)

nodes_g$tomb <- droplevels(factor(nodes_g$tomb))

# Create graph
g_g <- graph_from_data_frame(d = edges_g, vertices = nodes_g, directed = FALSE)

# Plot
ggraph(g_g, layout = "fr") +
  geom_edge_link(color = "grey70") +
  geom_node_point(aes(color = tomb), size = 3) +
  geom_node_text(aes(label = name), vjust = -1, size = 2) +
  theme_void() +
  ggtitle("All tombs with grouped Tomb 46 56 62") +
  scale_color_brewer(palette = "Set2")
```

# With kinship degrees

```{r, echo=FALSE}
network_grouped <- kinship_result %>%
  # group the 46/56/62 tombs
  mutate(
    tomb1 = ifelse(tomb1 %in% grouped_tombs, group_name, tomb1),
    tomb2 = ifelse(tomb2 %in% grouped_tombs, group_name, tomb2)
  ) %>%
  # keep only true kinship relationships
  filter(rel %in% c("First Degree", "IdenticalTwins/SameIndividual", "Second Degree", "Third Degree")) %>%
  mutate(
    # Manual edge widths for plotting
    edge_width = case_when(
      rel %in% c("First Degree", "IdenticalTwins/SameIndividual") ~ 2.0,
      rel == "Second Degree" ~ 1.2,
      rel == "Third Degree" ~ 0.6
    ),
    # Factor for legend ordering
    degree_label = factor(rel,
                          levels = c("First Degree", "Second Degree", "Third Degree"),
                          labels = c("1st Degree", "2nd Degree", "3rd Degree"))
  )

# --- Create edges and nodes ---

edges_g <- network_grouped %>%
  select(individual1_id, individual2_id, tomb1, tomb2, edge_width, degree_label)

nodes_g <- edges_g %>%
  transmute(name = individual1_id, tomb = tomb1) %>%
  bind_rows(edges_g %>% transmute(name = individual2_id, tomb = tomb2)) %>%
  distinct(name, .keep_all = TRUE)

nodes_g$tomb <- droplevels(factor(nodes_g$tomb))

# --- Build graph ---

g_g <- graph_from_data_frame(d = edges_g, vertices = nodes_g, directed = FALSE)

# --- Plot ---

set.seed(123)
ggraph(g_g, layout = "fr") +
  geom_edge_link(aes(width = edge_width, color = degree_label)) +
  geom_node_point(aes(color = tomb), size = 3) +
  geom_node_text(aes(label = name), vjust = -1, size = 2) +
  scale_edge_width(range = c(0.5, 2), guide = "none") +
  scale_edge_color_manual(
    values = c("1st Degree" = "black", "2nd Degree" = "grey40", "3rd Degree" = "grey70"),
    name = "Kinship degree"
  ) +
  scale_color_brewer(palette = "Set2") +
  ggtitle("All tombs with grouped Tomb 46 56 62") +
  theme_void()
```

#### Tomb 46 56 62 only

```{r only, echo=FALSE}
group_name <- "Tomb 46 56 62"

network_grouped <- kinship_result_grouped %>%
  # keep only pairs where at least one side is the grouped tomb
  filter(tomb1 == group_name | tomb2 == group_name) %>%
  
  mutate(
    related = case_when(
      rel %in% c("First Degree", "IdenticalTwins/SameIndividual", 
                 "Second Degree", "Third Degree") ~ TRUE,
      rel == "Uncertain" ~ FALSE,
      TRUE ~ NA
    )
  ) %>%
  filter(related == TRUE)

edges_g <- network_grouped %>%
  select(individual1_id, individual2_id, tomb1, tomb2)

nodes_g <- edges_g %>%
  transmute(name = individual1_id, tomb = tomb1) %>%
  bind_rows(edges_g %>% transmute(name = individual2_id, tomb = tomb2)) %>%
  distinct(name, .keep_all = TRUE)

nodes_g$tomb <- droplevels(factor(nodes_g$tomb))

g_g <- graph_from_data_frame(d = edges_g, vertices = nodes_g, directed = FALSE)

ggraph(g_g, layout = "fr") +
  geom_edge_link(color = "grey70") +
  geom_node_point(aes(color = tomb), size = 3) +
  geom_node_text(aes(label = name), vjust = -1, size = 2) +
  theme_void() +
  ggtitle("Network: Only pairs involving grouped Tomb 46 56 62") +
  scale_color_brewer(palette = "Set2")


```

# With kinship degrees

```{r, echo=FALSE}
group_name <- "Tomb 46 56 62"

network_grouped <- kinship_result_grouped %>%
  # Keep only pairs where at least one individual is in the grouped tomb
  filter(tomb1 == group_name | tomb2 == group_name) %>%
  # Keep only true kinship relationships
  filter(rel %in% c("First Degree", "IdenticalTwins/SameIndividual", "Second Degree", "Third Degree")) %>%
  mutate(
    # Manual edge widths
    edge_width = case_when(
      rel %in% c("First Degree", "IdenticalTwins/SameIndividual") ~ 2.0,
      rel == "Second Degree" ~ 1.2,
      rel == "Third Degree" ~ 0.6
    ),
    # Factor for legend ordering
    degree_label = factor(rel,
                          levels = c("First Degree", "Second Degree", "Third Degree"),
                          labels = c("1st Degree", "2nd Degree", "3rd Degree"))
  )

# --- Create edges and nodes ---
edges_g <- network_grouped %>%
  select(individual1_id, individual2_id, tomb1, tomb2, edge_width, degree_label)

nodes_g <- edges_g %>%
  transmute(name = individual1_id, tomb = tomb1) %>%
  bind_rows(edges_g %>% transmute(name = individual2_id, tomb = tomb2)) %>%
  distinct(name, .keep_all = TRUE)

nodes_g$tomb <- droplevels(factor(nodes_g$tomb))

# --- Build graph ---
g_g <- graph_from_data_frame(d = edges_g, vertices = nodes_g, directed = FALSE)

# --- Plot ---
set.seed(123)
ggraph(g_g, layout = "fr") +
  geom_edge_link(aes(width = edge_width, color = degree_label)) +
  geom_node_point(aes(color = tomb), size = 3) +
  geom_node_text(aes(label = name), vjust = -1, size = 2) +
  scale_edge_width(range = c(0.5, 2), guide = "none") +
  scale_edge_color_manual(
    values = c("1st Degree" = "black", "2nd Degree" = "grey40", "3rd Degree" = "grey70"),
    name = "Kinship degree"
  ) +
  scale_color_brewer(palette = "Set2") +
  ggtitle("Network: Only pairs involving grouped Tomb 46 56 62") +
  theme_void()
```

### Cross and within tomb relations

```{r, echo=FALSE}
# --- Prepare kinship network with weights for plotting ---
network_kr <- kinship_result %>%
  filter(rel %in% c("First Degree", "IdenticalTwins/SameIndividual", 
                    "Second Degree", "Third Degree")) %>%
  mutate(
    edge_width = case_when(
      rel %in% c("First Degree", "IdenticalTwins/SameIndividual") ~ 2.0,
      rel == "Second Degree" ~ 1.2,
      rel == "Third Degree" ~ 0.6
    ),
    degree_label = factor(rel,
                          levels = c("First Degree", "Second Degree", "Third Degree"),
                          labels = c("1st Degree", "2nd Degree", "3rd Degree"))
  )

# --- Create edges ---
edges <- network_kr %>%
  select(individual1_id, individual2_id, tomb1, tomb2, edge_width, degree_label) %>%
  mutate(
    edge_type = ifelse(tomb1 == tomb2, "Within-tomb", "Cross-tomb")
  )

# --- Create nodes ---
nodes <- edges %>%
  transmute(name = individual1_id, tomb = tomb1) %>%
  bind_rows(edges %>% transmute(name = individual2_id, tomb = tomb2)) %>%
  distinct(name, .keep_all = TRUE)

# --- Add sex information ---
nodes <- nodes %>%
  left_join(individual_metadata %>% select(individual_id, sex),
            by = c("name" = "individual_id"))

# Tomb factor
nodes$tomb <- droplevels(factor(nodes$tomb))

# --- List of tombs ---
tombs <- unique(nodes$tomb)

plots <- lapply(tombs, function(t) {
  
  edges_t <- edges %>% filter(tomb1 == t | tomb2 == t)
  
  nodes_t <- nodes %>%
    filter(name %in% unique(c(edges_t$individual1_id, edges_t$individual2_id))) %>%
    distinct(name, .keep_all = TRUE)    # <<< make vertex names unique
  
  # Build graph
  g_t <- graph_from_data_frame(d = edges_t, vertices = nodes_t, directed = FALSE)
  
  # Plot
  ggraph(g_t, layout = "fr") +
    geom_edge_link(aes(width = edge_width, color = edge_type), show.legend = TRUE) +
    geom_node_point(aes(color = tomb, shape = sex), size = 3) +
    geom_node_text(aes(label = name), vjust = -1, size = 2) +
    scale_edge_width(range = c(0.5, 2), guide = "none") +
    scale_edge_color_manual(values = c("Within-tomb" = "black", "Cross-tomb" = "red"),
                            name = "Edge Type") +
    scale_color_brewer(palette = "Set2") +
    scale_shape_manual(
      values = c("XX" = 16, "XY" = 17),
      labels = c("XX" = "Female", "XY" = "Male"),
      name = "Sex") +
    ggtitle(paste("Tomb:", t)) +
    theme_void()
})


# --- Print all plots ---
for (p in plots) print(p)

# Save each plot with the tomb name inside network_plots
# for (i in seq_along(plots)) {
#   tomb_name <- tombs[i]  # tombs is your vector of tomb names
#   ggsave(
#     filename = paste0("network_plots/", tomb_name, ".png"),
#     plot = plots[[i]],
#     width = 6, height = 4, dpi = 300
#   )
# }
```

# Grouped tomb 46 56 62 version
```{r, echo=FALSE}
# --- Group definition ---
grouped_tombs <- c("Elateia T46", "Elateia T56", "Elateia T62")
group_label   <- "Elateia T46 56 62"

# --- Prepare kinship network with grouping ---
network_kr <- kinship_result %>%
  mutate(
    tomb1 = ifelse(tomb1 %in% grouped_tombs, group_label, tomb1),
    tomb2 = ifelse(tomb2 %in% grouped_tombs, group_label, tomb2)
  ) %>%
  filter(rel %in% c(
    "First Degree",
    "IdenticalTwins/SameIndividual",
    "Second Degree",
    "Third Degree"
  )) %>%
  mutate(
    edge_width = case_when(
      rel %in% c("First Degree", "IdenticalTwins/SameIndividual") ~ 2.0,
      rel == "Second Degree" ~ 1.2,
      rel == "Third Degree" ~ 0.6
    ),
    degree_label = factor(
      rel,
      levels = c("First Degree", "Second Degree", "Third Degree"),
      labels = c("1st Degree", "2nd Degree", "3rd Degree")
    )
  )

# --- Create edges ---
edges <- network_kr %>%
  select(individual1_id, individual2_id, tomb1, tomb2, edge_width, degree_label) %>%
  mutate(
    edge_type = ifelse(tomb1 == tomb2, "Within-tomb", "Cross-tomb")
  )

# --- Create nodes ---
nodes <- edges %>%
  transmute(name = individual1_id, tomb = tomb1) %>%
  bind_rows(edges %>% transmute(name = individual2_id, tomb = tomb2)) %>%
  distinct(name, .keep_all = TRUE)

# --- Add sex information ---
nodes <- nodes %>%
  left_join(
    individual_metadata %>% select(individual_id, sex),
    by = c("name" = "individual_id")
  )

nodes$tomb <- droplevels(factor(nodes$tomb))

# --- Plot per tomb (grouped tomb included) ---
tombs <- unique(nodes$tomb)

plots <- lapply(tombs, function(t) {

  edges_t <- edges %>% filter(tomb1 == t | tomb2 == t)

  nodes_t <- nodes %>%
    filter(name %in% unique(c(edges_t$individual1_id, edges_t$individual2_id))) %>%
    distinct(name, .keep_all = TRUE)

  g_t <- igraph::graph_from_data_frame(
    d = edges_t,
    vertices = nodes_t,
    directed = FALSE
  )

  ggraph(g_t, layout = "fr") +
    geom_edge_link(aes(width = edge_width, color = edge_type)) +
    geom_node_point(aes(color = tomb, shape = sex), size = 3) +
    geom_node_text(aes(label = name), vjust = -1, size = 2) +
    scale_edge_width(range = c(0.5, 2), guide = "none") +
    scale_edge_color_manual(
      values = c("Within-tomb" = "black", "Cross-tomb" = "red"),
      name = "Edge Type"
    ) +
    scale_color_brewer(palette = "Set2") +
    scale_shape_manual(
      values = c("XX" = 16, "XY" = 17),
      labels = c("XX" = "Female", "XY" = "Male"),
      name = "Sex"
    ) +
    ggtitle(paste("Tomb:", t)) +
    theme_void()
})

# --- Print all plots ---
for (p in plots) print(p)

# # --- Save only the grouped tomb plot ---
# group_plot <- plots[[which(tombs == "Elateia T46 56 62")]]
# 
# ggsave(
#   filename = "network_plots/Tomb 46 56 62.png",
#   plot = group_plot,
#   width = 6,
#   height = 4,
#   dpi = 300
# )


```

**Comment:**

-   Different ratio in whether the related pairs or in the same tombs or in different tombs
-   But just by looking, some tombs really point out that they tends to have internal relatedness more than random relatedness to different tombs.
