---
title: "Marginal effect model"
author: "Felicia Francis Immanuel"
date: "2025-12-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
source(here::here("environment_setup.R"))
library(marginaleffects)
library(collapse)

source(here::here("Analysis", "pairs_data.R"))
imp_binary_group_cat <- readRDS(
  here::here("unofficial work", "models save", "imp_binary_group_cat.rds")
)
imp_binary_separate_cat <- readRDS(
  here::here("unofficial work", "models save", "imp_binary_separate_cat.rds")
)
```

## Interpreting GLMM coefficients with marginal effects

Our outcome is binary (`related = 1` vs `related = 0`), so we fit a logistic GLMM.  
In a logistic model, coefficients are on the log-odds scale:

\[
\text{logit}(p) = \log\left(\frac{p}{1-p}\right) = \eta = \beta_0 + \beta_1 x_1 + \dots
\]

Exponentiating a coefficient gives an odds ratio:

\[
\exp(\beta_k) = \text{odds ratio}
\]

This is not the same as “times more likely” in terms of probability.  
The same odds ratio can correspond to very different probability changes depending on the baseline probability \(p\). Therefore, interpreting \(\exp(\beta)\) as a direct multiplier of probability can be misleading when baseline probabilities are small.

## Marginal Effect

A marginal effect translates the model from the log-odds scale to the probability scale.  
It answers the question:

> “Holding other variables constant, how much does the predicted probability \(P(\text{related}=1)\) change when a predictor changes?”

For a continuous predictor \(x\), the (instantaneous) marginal effect is:

\[
\frac{\partial p}{\partial x}
\]

For a categorical predictor (e.g., sex-pair group), the marginal effect is typically reported as a discrete change:

\[
\Delta p = p(x=\text{level A}) - p(x=\text{reference})
\]

## Average marginal effects and marginal predicted probabilities

In practice, we report **average marginal effects (AMEs)**, which average the probability changes across individuals/pairs in the dataset. This yields results on an interpretable scale (percentage-point changes):

- **Marginal predicted probabilities**: predicted \(p\) for each group (e.g., XX–XX, XX–XY, XY–XY)
- **Average marginal effects**: average difference in predicted probability between groups (e.g., XY–XY vs XX–XX)

**What is the average probability that a pair is related, for: **

1. each sex_pair category?
```{r}
pp_group_sex <- avg_predictions(
  model = imp_binary_group_cat,
  by = "sex_pair",
  type = "response",# probability scale
  newdata = pairs_group, 
  re_formula = NA, # population-level (no random effects)
  allow_new_levels = TRUE
)

pp_group_sex

```
```{r}
ggplot(pp_group_sex, aes(x = sex_pair, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.1
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1)
  ) +
  labs(
    x = "Sex pair",
    y = "Predicted probability of relatedness",
    title = "Marginal predicted probabilities by sex pair",
    subtitle = "Population-level predictions (random effects excluded)"
  ) +
  theme_minimal()
```

2. each age_pair category
```{r}
pp_group_age <- avg_predictions(
  model = imp_binary_group_cat,
  by = "age_pair",
  type = "response",        
  newdata = pairs_group,
  re_formula = NA,          
  allow_new_levels = TRUE
)

pp_group_age

```
```{r}
ggplot(pp_group_age, aes(x = age_pair, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.1
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1)
  ) +
  labs(
    x = "Age pair",
    y = "Predicted probability of relatedness",
    title = "Marginal predicted probabilities by age pair",
    subtitle = "Population-level predictions (random effects excluded)"
  ) +
  theme_minimal()
```

3. each bone_quality category
```{r}
pp_group_bone <- avg_predictions(
  model = imp_binary_group_cat,
  by = "bone_quality",
  type = "response",        
  newdata = pairs_group,
  re_formula = NA,          
  allow_new_levels = TRUE
)

pp_group_bone

```
```{r}
ggplot(pp_group_bone, aes(x = bone_quality, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.1
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1)
  ) +
  labs(
    x = "Bone quality",
    y = "Predicted probability of relatedness",
    title = "Marginal predicted probabilities by bone quality",
    subtitle = "Population-level predictions (random effects excluded)"
  ) +
  theme_minimal()
```

4. each same_tomb category?
```{r}
pp_group_tomb <- avg_predictions(
  model = imp_binary_group_cat,
  by = "same_tomb",
  type = "response",        
  newdata = pairs_group,
  re_formula = NA,          
  allow_new_levels = TRUE
)

pp_group_tomb

```
```{r}
ggplot(pp_group_tomb, aes(x = same_tomb, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.1
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1)
  ) +
  labs(
    x = "Same tomb",
    y = "Predicted probability of relatedness",
    title = "Marginal predicted probabilities by tomb",
    subtitle = "Population-level predictions (random effects excluded)"
  ) +
  theme_minimal()
```

**“By how many percentage points does the probability change compared to the reference?”**

```{r}
new_data <- imp_binary_group_cat$data
```

1. sex pair
```{r}
cmp_group_sex <- avg_comparisons(
  imp_binary_group_cat,
  variables = "sex_pair",
  type = "response",
  newdata = new_data,
  re_formula = NA,
  allow_new_levels = TRUE
)
cmp_group_sex

```

2. age pair
```{r}
cmp_group_age <- avg_comparisons(
  imp_binary_group_cat,
  variables = "age_pair",
  type = "response",
  newdata = new_data,
  re_formula = NA,
  allow_new_levels = TRUE
)
cmp_group_age

```

3. bone quality

```{r}
cmp_group_bone <- avg_comparisons(
  imp_binary_group_cat,
  variables = "bone_quality",
  type = "response",
  newdata = new_data,
  re_formula = NA,
  allow_new_levels = TRUE
)
cmp_group_bone

```


4. same tomb

```{r}
cmp_group_tomb <- avg_comparisons(
  imp_binary_group_cat,
  variables = "same_tomb",
  type = "response",
  newdata = new_data,
  re_formula = NA,
  allow_new_levels = TRUE
)
cmp_group_tomb

```