---
title: "Model compare"
output: html_document
---

``` {r setup, include=FALSE}
library(here)
source(here("environmentSetUp.R"))
imp_binary_separate_cat <- readRDS(here::here("unofficial work/models save/imp_binary_separate_cat.rds"))
all_binary_separate_cat <- readRDS(here::here("unofficial work/models save/all_binary_separate_cat.rds"))
imp_binary_separate_cat <- readRDS(here::here("unofficial work/models save/imp_binary_separate_cat.rds"))
imp_binary_separate_nobone <- readRDS(here::here("unofficial work/models save/imp_binary_separate_nobone.rds"))
imp_binary_separate_newtombvar <- readRDS(here::here("unofficial work/models save/imp_binary_separate_newtombvar.rds"))
imp_binary_separate_withintomb <- readRDS(here("unofficial work/models save/imp_binary_separate_withintomb.rds"))
imp_binary_separate_unrel3rd <- readRDS(here("unofficial work/models save/imp_binary_separate_unrel3rd.rds"))
imp_binary_separate_sametomb <- readRDS(here("unofficial work/models save/imp_binary_separate_sametomb.rds"))
imp_binary_separate_nobone_prior <- readRDS(here("unofficial work/models save/imp_binary_separate_nobone_prior.rds"))
imp_binary_separate_withintomb_unrel3rd <- readRDS(here("unofficial work/models save/imp_binary_separate_withintomb_unrel3rd.rds"))

imp_binary_group_cat <- readRDS(here::here("unofficial work/models save/imp_binary_group_cat.rds"))
all_binary_group_cat <- readRDS(here::here("unofficial work/models save/all_binary_group_cat.rds"))
imp_binary_group_cat <- readRDS(here::here("unofficial work/models save/imp_binary_group_cat.rds"))
imp_binary_group_nobone <- readRDS(here::here("unofficial work/models save/imp_binary_group_nobone.rds"))
imp_binary_group_newtombvar <- readRDS(here::here("unofficial work/models save/imp_binary_group_newtombvar.rds"))
imp_binary_group_withintomb <- readRDS(here("unofficial work/models save/imp_binary_group_withintomb.rds"))
imp_binary_group_unrel3rd <- readRDS(here("unofficial work/models save/imp_binary_group_unrel3rd.rds"))
imp_binary_group_sametomb <- readRDS(here("unofficial work/models save/imp_binary_group_sametomb.rds"))
imp_binary_group_nobone_prior <- readRDS(here("unofficial work/models save/imp_binary_group_nobone_prior.rds"))
imp_binary_group_withintomb_unrel3rd <- readRDS(here("unofficial work/models save/imp_binary_group_withintomb_unrel3rd.rds"))

## Define compare function
compare_model <- function(
  models,
  ci_width = 0.95,
  digits = 2,
  output = c("gt", "data"),
  title = "Model comparison"
) {
  output <- match.arg(output)
  stopifnot(is.list(models), !is.null(names(models)))

  extract_one <- function(fit, model_name) {
    tidybayes::tidy_draws(fit) |>
      tidyr::pivot_longer(
        cols = dplyr::matches("^(b_|sd_)"),
        names_to = "term",
        values_to = "value"
      ) |>
      dplyr::group_by(term) |>
      tidybayes::median_qi(value, .width = ci_width) |>
      dplyr::transmute(
        term,
        model = model_name,
        estimate = value,
        ci_low = .lower,
        ci_high = .upper
      )
  }

  long <- purrr::imap_dfr(models, extract_one)

  wide <- long |>
    dplyr::mutate(
      term = dplyr::case_when(
        term == "b_Intercept" ~ "Intercept",
        grepl("^b_", term) ~ gsub("^b_", "", term),
        grepl("^sd_", term) ~ gsub("^sd_", "SD(RE): ", term),
        TRUE ~ term
      ),
      CI = sprintf(
        paste0("[%.", digits, "f, %.", digits, "f]"),
        ci_low, ci_high
      ),
      row_order = dplyr::case_when(
        term == "Intercept" ~ 1,
        grepl("^SD\\(RE\\):", term) ~ 3,
        TRUE ~ 2
      )
    ) |>
    dplyr::select(term, model, estimate, CI, row_order) |>
    tidyr::pivot_wider(
      names_from = model,
      values_from = c(estimate, CI),
      names_glue = "{model}_{.value}"
    )

  col_order <- c(
    "term",
    unlist(
      lapply(
        names(models),
        function(m) c(paste0(m, "_estimate"), paste0(m, "_CI"))
      )
    )
  )

  tab <- wide |>
    dplyr::arrange(row_order, term) |>
    dplyr::select(dplyr::all_of(col_order))

  # ---- default: return gt table ----
  if (output == "gt") {
    gt_tab <- gt::gt(tab) |>
      gt::tab_header(
        title = title
      ) |>
      gt::cols_align(align = "left", term) |>
      gt::cols_align(align = "center", -term)

    # automatic column spanners per model
    for (m in names(models)) {
      gt_tab <- gt_tab |>
        gt::tab_spanner(
          label = m,
          columns = c(paste0(m, "_estimate"), paste0(m, "_CI"))
        )
    }

    return(gt_tab)
  }

  return(tab)
}


```

### Remark
Every models here are fitted with 15k threshold. Except the one was used to compare 2  threshold.

### Compare between with and without bone quality
``` {r compare bone}
summary(imp_binary_separate_cat)
summary(imp_binary_separate_nobone)
compare_bone <- list("with bone" = imp_binary_separate_cat,
                     "no bone" = imp_binary_separate_nobone)
compare_model(compare_bone)
```

**Comment:**

- The model stays stable, estimates and CI change minimally
- No bone model even gives better diagnostics $\Rightarrow$ mathematically cleaner
- $\Rightarrow$ Prefer the no bone model, since it's simpler, easier to interpret, and mathematically cleaner without losing scientific accuracy (also it's theoretically justified by DAG).

### Compare between 15k threshold and original threhold
#### Separated tomb
``` {r compare threshold}
# 15k threshold
summary(imp_binary_separate_cat)

# Standard threshold
summary(all_binary_separate_cat)

stdthreshold_vs_15k <- list("15k threshold" = imp_binary_separate_cat,
                            "standard threshold" = all_binary_separate_cat)
compare_model(stdthreshold_vs_15k)
```

**Comment:**

- Model with original threshold: high individual variance, mostly larger CI
- However most of the estimates give the same qualitative results (except b_bone_qualityhighMlow but we have decided not to include bone quality already)


### Add tomb variables
Additional tomb-level variables: sample_success_min_norm, analysed_ind_min_norm, 
length_of_use_average_norm

``` {r tomb var}
imp_binary_separate_newtombvar

normal_vs_tombvar <- list(
  "Normal (no bone)" = imp_binary_separate_nobone,
  "With tomb variables" = imp_binary_separate_newtombvar
)

compare_model(normal_vs_tombvar)
```

- Tomb variables have large estimates with huge CI
- Old parameters mostly stay stable
- Parameter for sample_success_min is strong NEGATIVE:
  - Interpretation: Higher success rate => Lower probability of relatedness
  - **How this could make sense:** 
    - High success rate => analyse many random people from the tomb => higher unrelated pairs (since we have low related rate) => family members are diluted in large # random people
    - Low success rate => only few bones survive the poor preservation context => those who survive might be clustered in a specific protected spot => could be a spot that is preserved for family => higher probability of relatedness
  - Anyway this parameter is added to control this "diluted bias" more than to describe any reality

### Robust check
Now we do robust check with:

- Only same_tomb as predictor
- A data set with only "within tomb relationship"
- New definition of relatedness: 3rd degree is categorized as unrelated

``` {r robust}
imp_binary_separate_withintomb
imp_binary_separate_unrel3rd

robustcheck_models <- list(
  "Normal (no bone)" = imp_binary_separate_nobone,
  "Only same_tomb" = imp_binary_separate_sametomb,
  "Within tomb" = imp_binary_separate_withintomb,
  "3rd as unrelated" = imp_binary_separate_unrel3rd
)

compare_model(robustcheck_models)
```

**Comment:**

- **Context:** same_tomb is a very dominant predictor, we do robust check to see if the other variables are just artifacts of this same_tomb
- The compare table shows that all estimates have the same qualitative result. 
- All the estimates decrease toward 0. However MalexMale effect stays pretty strong through all the models
- **Remark for "within tomb" model:** This model is only fitted with 2990 observations (compare to 13690 for all other models). So maybe the power is not as high.
- **"3rd as unrelated" model:** same_tomb gets much higher. $\Rightarrow$ Share burial is a strong predictor for immediate kinship? (Intuition: 3rd degree is further kin, they might not live in the same immediate household or be buried in the exact same family crypt as strictly as a parent and child would. By including them in "Related", we dilute this effect (?))



### Compare everything (separated tomb version)
``` {r everything}
every_model <- list(
  "Normal (with bone)" = imp_binary_separate_cat,
  "Normal (no bone)" = imp_binary_separate_nobone,
  "With tomb variables" = imp_binary_separate_newtombvar,
  "Only same_tomb" = imp_binary_separate_sametomb,
  "Within tomb" = imp_binary_separate_withintomb,
  "3rd as unrelated" = imp_binary_separate_unrel3rd
)

compare_model(every_model, title = "Model comparison (separated tombs)")
```


### Compare everything (grouped tomb version) 
``` {r everything group}
every_model_group <- list(
  "G Normal (with bone)" = imp_binary_group_cat,
  "G Normal (no bone)" = imp_binary_group_nobone,
  "G With tomb variables" = imp_binary_group_newtombvar,
  "G Only same_tomb" = imp_binary_group_sametomb,
  "G Within tomb" = imp_binary_group_withintomb,
  "G 3rd as unrelated" = imp_binary_group_unrel3rd
)

compare_model(every_model_group, title = "Model comparison (grouped tombs)")
```

**Comment:**

- "G With tomb variables" model didn't converge. Possible reason: grouping tombs reduced the number of unique "tomb environments."
- The other estimates stay stable along the grouped models and behave quite similar to the "separated version"
- same_tomb is always stronger than in the corresponding separated models (since for larger group of tomb we can expect to find more related pairs). 
  - Especially same_tomb rockets in the "Within tomb" model


### Grouped version models' output
``` {r group outputs}
imp_binary_group_cat
imp_binary_group_nobone
imp_binary_group_newtombvar
imp_binary_group_sametomb
imp_binary_group_withintomb
imp_binary_group_unrel3rd
```


### Compare between flat prior and weak informative prior
- Sensitivity analysis to see if flat prior is justified. Or if our previous model was just "noise chasing". 
- Now we use Weakly Informative Prior (Normal(0, 1)) to pulls estimate toward zero unless the data strongly disagrees.
- If the flat prior was wrong: The coefficients in the Normal(0,1) model would have collapsed to zero or changed direction, proving the original results were just noise.

``` {r prior}
imp_binary_separate_nobone_prior
compare_model(list("Main model with flat prior" = imp_binary_separate_nobone,
                   "Main model with N(0,1) prior" = imp_binary_separate_nobone_prior))

```

**Comments:**

- The coefficients shrank toward 0 (as expected because of N(0,1) regularization)
- However they didn't change the direction
- same_tomb remains the dominant predictor and barely affected by the prior

### Try out the "strict family in internal structure" model
- This model is fitted with the pairs in the same tomb only (within tomb relationship), and with the stricter definition of relatedness (1st/2nd as related, 3rd as unrelated)
- Equivalent to a pure test for biological predictors like sex and age

``` {r strict}
imp_binary_separate_withintomb_unrel3rd

compare_model(list(
  "Within tomb" = imp_binary_separate_withintomb,
  "3rd as unrelated" = imp_binary_separate_unrel3rd,
  "Within tomb and 3rd as unrelated" = imp_binary_separate_withintomb_unrel3rd
))

```

**Comments:**

- All the coefficients shrunk strongly towards 0
- Most of the CIs cross 0 now, maybe there is no biological structure in this setting??
- Possible additional reason: power issue (only 2990 pairs), "rare event" (too few related pairs within one tomb, and now we even reduce the related pairs by couting 3rd degree as unrelated)

















