---
title: "new AD model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(loo)
source(here::here("environmentSetUp.R"))
source(here::here("Analysis/pairsdata.R"))
```

**Remark:**

- Only model the pairs over 15k in Overlapping SNPs
- Create new variable to collapse skeletal elements into good and bad bones only (high-quality: petrous, low-quality: others)
- Use beta distrbution for numeric target (kinship coefficient) since it's skewed heavily

### Separated tomb
#### Binary separated (with categorical bones)
``` {r bin sep cat}
binary_separate_cat <- readRDS(here::here("unofficial work/models save/binary_separate_cat.rds"))

summary(binary_separate_cat)
```

#### Binary separated (with Overlapping SNPs instead of bones)
``` {r bin sep ov}
binary_separate_ov <- readRDS(here::here("unofficial work/models save/binary_separate_ov.rds"))

summary(binary_separate_ov)

```

#### Numeric separated (with categorical bones)
``` {r num sep cat}
numeric_separate_cat <- readRDS(here::here("unofficial work/models save/numeric_separate_cat.rds"))

summary(numeric_separate_cat)
```

#### Numeric separated (with Overlapping SNPs instead of bones)
``` {r num sep ov}
numeric_separate_ov <- readRDS(here::here("unofficial work/models save/numeric_separate_ov.rds"))

summary(numeric_separate_ov)

```


### Compare models
#### Model health
``` {r compare}
# 1. Put your models in a list (Make sure these objects exist in your environment)
model_list <- list(
  "binary_separate_cat" = binary_separate_cat, 
  "binary_separate_ov"  = binary_separate_ov,
  "numeric_separate_cat" = numeric_separate_cat, # Be careful if this one is broken/missing
  "numeric_separate_ov"  = numeric_separate_ov   # Be careful if this one is broken/missing
)

# 2. Corrected Function
get_model_stats <- function(model, name) {
  
  # A. Model Health
  # Use nuts_params from brms to get divergences
  nuts <- tryCatch(nuts_params(model), error = function(e) NULL)
  if(!is.null(nuts)) {
    divs <- sum(subset(nuts, Parameter == "divergent__")$Value)
  } else {
    divs <- NA # Handle cases where model crashed
  }
  
  # FIX FOR RHAT ERROR:
  # We extract Rhats directly from the rhat() function in brms namespace
  # If that fails, we return NA
  r_values <- tryCatch(brms::rhat(model), error = function(e) NA)
  rhat_max <- max(r_values, na.rm = TRUE)
  
  # B. Model Fit (WAIC)
  waic_val <- tryCatch(waic(model)$estimates["waic", "Estimate"], error = function(e) NA)
  
  # C. Key Biological Coefficients
  # We use fixef() which is standard for brms
  fix <- fixef(model)
  
  # Extract Male-Male Effect (Adjust string "XYMXY" if your summary shows differently)
  # using grep to find the row that matches "XYMXY" even if the name varies slightly
  mm_row <- grep("XYMXY", rownames(fix))
  beta_MM <- if(length(mm_row) > 0) fix[mm_row, "Estimate"] else NA
  
  # Extract Subadult Effect
  sub_row <- grep("subadult", rownames(fix))
  beta_Subadult <- if(length(sub_row) > 0) fix[sub_row, "Estimate"] else NA
  
  # D. Random Effect Variances
  # Safely extract variances
  vars <- VarCorr(model)
  
  # Extract Individual Variance (mm function)
  # Note: The name is usually 'mmindividual1_idindividual2_id' but might vary
  sd_ind <- NA
  if("mmindividual1_idindividual2_id" %in% names(vars)) {
    sd_ind <- vars$mmindividual1_idindividual2_id$sd[1]
  }
  
  # Extract Tomb Variance
  sd_tomb <- NA
  if("mmtomb1tomb2" %in% names(vars)) {
    sd_tomb <- vars$mmtomb1tomb2$sd[1]
  }
  
  return(data.frame(
    Model = name,
    Divergences = divs,
    Max_Rhat = round(rhat_max, 3),
    WAIC = round(waic_val, 1),
    Est_MaleMale = round(beta_MM, 2),
    Est_Subadult = round(beta_Subadult, 2),
    SD_Ind = round(sd_ind, 2),
    SD_Tomb = round(sd_tomb, 2)
  ))
}

# 3. Run and Bind
comparison_table <- bind_rows(lapply(names(model_list), function(x) get_model_stats(model_list[[x]], x)))

# 4. View
comparison_table
```


**Comment:**

- The healthiest model is binary_separated_cat
- Although there are still divergence (11, 0.275%) but we have good Rhat ($\approx 1$), 
and good Effective Sample Size (>1000)

**Source: **
- https://mc-stan.org/learn-stan/diagnostics-warnings.html
- https://easystats.github.io/bayestestR/reference/effective_sample.html

**Improve: **
- Rerun with higher adapt_delta (0.999) and more iteration

### Improved version
#### Separated tomb
``` {r imp sep}
imp_binary_separate_cat <- readRDS(here::here("unofficial work/models save/imp_binary_separate_cat.rds"))
summary(imp_binary_separate_cat)
```

#### Group tomb
``` {r imp gr}
imp_binary_group_cat <- readRDS(here::here("unofficial work/models save/imp_binary_group_cat.rds"))
summary(imp_binary_group_cat)
```

**Comment:**

- So much better outcome
- No divergence, Rhat = 1, all large ESS
- No drastical difference compare to the previous model with divergence, indicate that the 
11 divergences were negligible
- The "Same Tomb" effect got stronger 
  - If the grouping was random, the effect would likely be weaker
  - But it got stronger => the relatedness in grouped tomb is more confirmed
  

















