---
title: "new AD model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(loo)
source(here::here("environmentSetUp.R"))
source(here::here("Analysis/pairsdata.R"))
```

```{r plot}
plot(density(pairs_ind$nonnormalized_p0))
plot(density(pairs_ind$kinship_coefficient))
```

**Remark:**

- Only model the pairs over 15k in Overlapping SNPs
- Create new variable to collapse skeletal elements into good and bad bones only (high-quality: petrous, low-quality: others)
- Use beta distrbution for numeric target (kinship coefficient) since it's skewed heavily

### Separated tomb
#### Binary separated (with categorical bones)
``` {r bin sep cat}
binary_separate_cat <- readRDS(here::here("unofficial work/models save/binary_separate_cat.rds"))

summary(binary_separate_cat)
```

#### Binary separated (with Overlapping SNPs instead of bones)
``` {r bin sep ov}
binary_separate_ov <- readRDS(here::here("unofficial work/models save/binary_separate_ov.rds"))

summary(binary_separate_ov)

```

#### Numeric separated (with categorical bones)
``` {r num sep cat}
numeric_separate_cat <- readRDS(here::here("unofficial work/models save/numeric_separate_cat.rds"))

summary(numeric_separate_cat)
```

#### Numeric separated (with Overlapping SNPs instead of bones)
``` {r num sep ov}
numeric_separate_ov <- readRDS(here::here("unofficial work/models save/numeric_separate_ov.rds"))

summary(numeric_separate_ov)

```


### Compare models
#### Model health
``` {r compare}
# 1. Put your models in a list (Make sure these objects exist in your environment)
model_list <- list(
  "binary_separate_cat" = binary_separate_cat, 
  "binary_separate_ov"  = binary_separate_ov,
  "numeric_separate_cat" = numeric_separate_cat, # Be careful if this one is broken/missing
  "numeric_separate_ov"  = numeric_separate_ov   # Be careful if this one is broken/missing
)

# 2. Corrected Function
get_model_stats <- function(model, name) {
  
  # A. Model Health
  # Use nuts_params from brms to get divergences
  nuts <- tryCatch(nuts_params(model), error = function(e) NULL)
  if(!is.null(nuts)) {
    divs <- sum(subset(nuts, Parameter == "divergent__")$Value)
  } else {
    divs <- NA # Handle cases where model crashed
  }
  
  # FIX FOR RHAT ERROR:
  # We extract Rhats directly from the rhat() function in brms namespace
  # If that fails, we return NA
  r_values <- tryCatch(brms::rhat(model), error = function(e) NA)
  rhat_max <- max(r_values, na.rm = TRUE)
  
  # B. Model Fit (WAIC)
  waic_val <- tryCatch(waic(model)$estimates["waic", "Estimate"], error = function(e) NA)
  
  # C. Key Biological Coefficients
  # We use fixef() which is standard for brms
  fix <- fixef(model)
  
  # Extract Male-Male Effect (Adjust string "XYMXY" if your summary shows differently)
  # using grep to find the row that matches "XYMXY" even if the name varies slightly
  mm_row <- grep("XYMXY", rownames(fix))
  beta_MM <- if(length(mm_row) > 0) fix[mm_row, "Estimate"] else NA
  
  # Extract Subadult Effect
  sub_row <- grep("subadult", rownames(fix))
  beta_Subadult <- if(length(sub_row) > 0) fix[sub_row, "Estimate"] else NA
  
  # D. Random Effect Variances
  # Safely extract variances
  vars <- VarCorr(model)
  
  # Extract Individual Variance (mm function)
  # Note: The name is usually 'mmindividual1_idindividual2_id' but might vary
  sd_ind <- NA
  if("mmindividual1_idindividual2_id" %in% names(vars)) {
    sd_ind <- vars$mmindividual1_idindividual2_id$sd[1]
  }
  
  # Extract Tomb Variance
  sd_tomb <- NA
  if("mmtomb1tomb2" %in% names(vars)) {
    sd_tomb <- vars$mmtomb1tomb2$sd[1]
  }
  
  return(data.frame(
    Model = name,
    Divergences = divs,
    Max_Rhat = round(rhat_max, 3),
    WAIC = round(waic_val, 1),
    Est_MaleMale = round(beta_MM, 2),
    Est_Subadult = round(beta_Subadult, 2),
    SD_Ind = round(sd_ind, 2),
    SD_Tomb = round(sd_tomb, 2)
  ))
}

# 3. Run and Bind
comparison_table <- bind_rows(lapply(names(model_list), function(x) get_model_stats(model_list[[x]], x)))

# 4. View
comparison_table
```


**Comment:**

- The healthiest model is binary_separated_cat
- Although there are still divergence (11, 0.275%) but we have good Rhat ($\approx 1$), 
and good Effective Sample Size (>1000)

**Source: **
- https://mc-stan.org/learn-stan/diagnostics-warnings.html
- https://easystats.github.io/bayestestR/reference/effective_sample.html

**Improve: **
- Rerun with higher adapt_delta (0.999) and more iteration

### Improved version
#### Separated tomb
``` {r imp sep}
imp_binary_separate_cat <- readRDS(here::here("unofficial work/models save/imp_binary_separate_cat.rds"))
summary(imp_binary_separate_cat)
```

#### Group tomb
``` {r imp gr}
imp_binary_group_cat <- readRDS(here::here("unofficial work/models save/imp_binary_group_cat.rds"))
summary(imp_binary_group_cat)
```

**Comment:**

- So much better outcome
- No divergence, Rhat = 1, all large ESS
- No drastical difference compare to the previous model with divergence, indicate that the 
11 divergences were negligible
- The "Same Tomb" effect got stronger 
  - If the grouping was random, the effect would likely be weaker
  - But it got stronger => the relatedness in grouped tomb is more confirmed
  


### Compare between 15k threshold and original threhold
#### Separated tomb
``` {r compare threshold}
imp_binary_separate_cat <- readRDS(here::here("unofficial work/models save/imp_binary_separate_cat.rds"))
all_binary_separate_cat <- readRDS(here::here("unofficial work/models save/all_binary_separate_cat.rds"))
ogthreshold_vs_15k <- readRDS(here::here("unofficial work/models save/ogthreshold_vs_15k.rds"))


summary(imp_binary_separate_cat)
summary(all_binary_separate_cat)
ogthreshold_vs_15k
```


**Comment:**

- Model with original threshold is distorted: high individual variance, most of the parameters are insignificant, worse diagnostics than the 15k threshold
- Comparison table point out that: 
  - Original threshold explains more variance (higher $R^2$)
  - 15k predicts new data better (ELPD, LOOIC, WAIC)
  - $\Rightarrow$ Indicates using original threshold is more complex or overfitting.



### Compare between with and without bone quality
Logic of this sensitivity analysis: 

Because "Tomb" $\rightarrow$ "Preservation" $\rightarrow$ "Observed kinship", therefore we added bone_quality to the model before to control for the path between "Tomb" and "Kinship". So that the relationship between Tomb and Kinship not just an artifact of preservation. 

But does this "preservation bias" really exist (and need to be controlled by the bone quality)? Or was it already taken care of from when we filter 15k threshold (only qualified sample)? 

$\Rightarrow$ Sensitivity analysis:

- Bias: If when we remove bone_quality, the "Same Tomb" coefficient shifts wildly because it suddenly has to absorb the preservation bias. Then we still need to add bone_quality in the model.
- No bias: When we remove bone_quality, the same_tomb coefficient stays stable. Then we can just delete bone_quality (which makes the model simpler and easier to interpret).

``` {r compare bone}
imp_binary_separate_cat <- readRDS(here::here("unofficial work/models save/imp_binary_separate_cat.rds"))
imp_binary_separate_nobone <- readRDS(here::here("unofficial work/models save/imp_binary_separate_nobone.rds"))
bone_vs_nobone <- readRDS(here::here("unofficial work/models save/bone_vs_nobone.rds"))


summary(imp_binary_separate_cat)
summary(imp_binary_separate_nobone)
bone_vs_nobone
```

**Comment:**
- The model stays stable
- No bone model even gives better diagnostics $\Rightarrow$ mathematically cleaner
- The compared metrics are nearly identical, no metric provides meaningful evidence that one model is better

$\Rightarrow$ Prefer the no bone model, since it's simpler, easier to interpret, and mathematically cleaner without losing scientific accuracy.




