---
title: "new AD model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
source(here::here("environmentSetUp.R"))
source(here::here("Analysis/pairsdata.R"))
```

**Remark:**

- Only model the pairs over 15k in Overlapping SNPs
- Create new variable to collapse skeletal elements into good and bad bones only (high-quality: petrous, low-quality: others)
- Use beta distrbution for numeric target (kinship coefficient) since it's skewed heavily

``` {r data}
pairs_ind <- pairs_ind[pairs_ind$overlap_nsnps >= 15000, ]
pairs_group <- pairs_group[pairs_group$overlap_nsnps >= 15000, ]
pairs_ind <- pairs_ind %>%
  mutate(
    bone_quality = case_when(
      grepl("petrous", skeletal_element1) & grepl("petrous", skeletal_element2) ~ "high-high",
      grepl("petrous", skeletal_element1) | grepl("petrous", skeletal_element2) ~ "high-low",
      TRUE ~ "low-low"
    )
  )
```

### Separated tomb
#### Binary separated (with categorical bones)
``` {r bin sep cat}
binary_separate_cat <- brm(related_binary ~ sex_pair + age_pair + bone_quality + same_tomb +
  (1 | mm(individual1_id, individual2_id)) +
  (1 | mm(tomb1, tomb2)),
  data = pairs_ind,
  family = bernoulli(),
  seed = 123)

summary(binary_separate_cat)
```

#### Binary separated (with Overlapping SNPs instead of bones)
``` {r bin sep ov}
binary_separate_ov <- brm(related_binary ~ sex_pair + age_pair + overlap_nsnps + same_tomb +
  (1 | mm(individual1_id, individual2_id)) +
  (1 | mm(tomb1, tomb2)),
  data = pairs_ind,
  family = bernoulli(),
  seed = 123)

summary(binary_separate_ov)

```

#### Numeric separated (with categorical bones)
``` {r num sep cat}
numeric_separate_cat <- brm(kinship_coefficient ~ sex_pair + age_pair + bone_quality + same_tomb +
  (1 | mm(individual1_id, individual2_id)) +
  (1 | mm(tomb1, tomb2)),
  data = pairs_ind,
  family = gaussian(),
  seed = 123)

summary(numeric_separate_cat)
```

#### Numeric separated (with Overlapping SNPs instead of bones)
``` {r num sep ov}
numeric_separate_ov <- brm(kinship_coefficient ~ sex_pair + age_pair + overlap_nsnps + same_tomb +
  (1 | mm(individual1_id, individual2_id)) +
  (1 | mm(tomb1, tomb2)),
  data = pairs_ind,
  family = gaussian(),
  seed = 123)

summary(numeric_separate_ov)

```


### Grouped tomb
#### Binary grouped (with categorical bones)
``` {r bin group cat}
binary_group_cat <- brm(related_binary ~ sex_pair + age_pair + bone_quality + same_tomb +
  (1 | mm(individual1_id, individual2_id)) +
  (1 | mm(tomb1, tomb2)),
  data = pairs_group,
  family = bernoulli(),
  seed = 123)

summary(binary_group_cat)
```

#### Binary grouped (with Overlapping SNPs instead of bones)
``` {r bin group ov}
binary_group_ov <- brm(related_binary ~ sex_pair + age_pair + overlap_nsnps + same_tomb +
  (1 | mm(individual1_id, individual2_id)) +
  (1 | mm(tomb1, tomb2)),
  data = pairs_group,
  family = bernoulli(),
  seed = 123)

summary(binary_group_ov)

```

#### Numeric grouped (with categorical bones)
``` {r num group cat}
numeric_group_cat <- brm(kinship_coefficient ~ sex_pair + age_pair + bone_quality + same_tomb +
  (1 | mm(individual1_id, individual2_id)) +
  (1 | mm(tomb1, tomb2)),
  data = pairs_group,
  family = gaussian(),
  seed = 123)

summary(numeric_group_cat)
```

#### Numeric grouped (with Overlapping SNPs instead of bones)
``` {r num group ov}
numeric_group_ov <- brm(kinship_coefficient ~ sex_pair + age_pair + overlap_nsnps + same_tomb +
  (1 | mm(individual1_id, individual2_id)) +
  (1 | mm(tomb1, tomb2)),
  data = pairs_group,
  family = gaussian(),
  seed = 123)

summary(numeric_group_ov)

```
